Recursive partitioning of linear growth curve models

### 2023-02-10 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
some improvements and comments in application
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3254 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 62318d166dfa1da296d90779a5c49a79c8f417c0

We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998-1999 \citep[ECLS-K; ][]{NCES10}. Data were collected from 21,304 children from 1,018 schools across the USA were collected. Assessments took place from kindergarten in 1998 through 8th grade in 2007. In the current analyses, we focus on the assessments from kindergarten, 1st, 3rd, 5th and 8th grade.

Response variables are reading, math and science abilities. These were assessed using multi-item cognitive tests, from which latent ability estimates were computed with a mean of zero and variance of one. Reading and math abilities were assessed in all five rounds of data collection, while science knowledge was assessed in the 3rd, 5th and 8th grade only. We analyzed data from children who completed all assessments yielding $N = 6,277$ for reading; $N = 6,512$ for math; $N = 6,625$ for science.

@@

We used eleven time-invariant covariates as potential partitioning variables: Gender (51.1\% male); Age in months at baseline (range 53 to 96; M = 6.14 years at baseline); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range -5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).

@@

@@

@@

The ECLS-K datasets have exceptionally large sample sizes. We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, likely more representative of real-world studies in psychology. For each combination of response variable (math, reading or science) and sample size, we performed 100 repetitions. We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition. Note that this measures the accuracy of the fixed-effects predictions only. Tree size was measured by the number of splits in each tree, and computation time was measured in seconds.

@@

## Ideally, would refit all longRPart_s trees which had convergence errors
## in one of the nodes. And then use NA for prediction, or refit a model
## without random slopes. This might paint a slightly difference picture.
## However, I do not expect dramatic differences, given that results
## for longRPart and longRPart_s are quite similar, and seem to differ
## most for Science, which was most strongly affected by the mistake.

@@

LongRPart trees comprising both random intercepts and slopes yielded estimation errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the Math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the Science data. All other model-fitting attempts converged successfully. Results are presented for successful model estimations.

@@

Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs. Only few differences in performance between LM(M) and longRPart trees are larger $> 1$ standard error: For Reading with $N=250$ and for Science with both sample sizes. Note that the standard errors in the pairwise $t$-tests are get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not. The latter thus provide a more useful measure of effect size, while the former have higher power to detect differences between methods.

Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting Math and Reading abilities. For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best. Random-effects initialization was the worst-performing LMM tree approach for Math and Reading.

For longRPart, trees comprising both a random intercept and slope performed best for predicting Math and Reading, trees with a random intercept only performed best for predicting Science. The differences in performance between Math and Reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of Science abilities.

@@

\begin{figure}[!bt]
\caption{Tree sizes for trees fitted to Math (top), Reading (middle) and Science (bottom) ability trajectories.}
\begin{subfigure}{1.2\textwidth}
<<echo=FALSE, fig=TRUE, height=2, width=8>>=
## Tree size
load("sizes, N = 250, Math.Rda")
math_sizes <- cbind(sizes, N = 250)
load("sizes, N = 1000, Math.Rda")
math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
#sapply(math_sizes, function(x) table(is.na(x)))
#sapply(math_sizes, function(x) tapply(x, math_sizes$N, mean, na.rm =TRUE))

@@

Figure~\ref{fig:application_sizes} presents tree size distributions for each of the partitioning approaches. Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees. For Math and Reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small. This pattern reverses with large sample size, where the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach. For the Math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.

>>>>

We analyzed longitudinal assessments of children's reading, math and science abilities from the Early Childhood Longitudinal Study-Kindergarten class of 1998--1999 \citep[ECLS-K; ][]{NCES10}. Data were collected from 21,304 children from 1,018 schools across the USA. Assessments took place from kindergarten in 1998 through 8th grade in 2007. In the current analyses, we focus on the assessments from kindergarten, 1st, 3rd, 5th and 8th grade.

Response variables are reading, math, and science abilities. These were assessed using multi-item cognitive tests, from which latent ability estimates were computed with a mean of zero and variance of one. Reading and math abilities were assessed in all five rounds of data collection, while science knowledge was assessed in the 3rd, 5th and 8th grade only. We analyzed data from children who completed all assessments yielding $N = 6,277$ for reading; $N = 6,512$ for math; $N = 6,625$ for science.

@@

\emph{README: Are these power transformations for months by 1/2 or 2/3 standard in the literature? Should this be cited in some way or is this uncontroversial?}

We used eleven time-invariant covariates as potential partitioning variables: Gender (51.1\% male); Age in months at baseline (range 53 to 96; M = 6.14 years at baseline); Race (8 categories); First time in kindergarten (yes/no); Socio-economic status (range $-$5 to 3); Fine motor skills (e.g., drawing figures; range 0 to 9); Gross motor skills (e.g., ability to hop, skip and jump; range 0 to 8); Interpersonal skills (range 1 to 4); Self-control (range 1 to 4); Internalizing problem behavior (range 1 to 4); Externalizing problem behavior (range 1 to 4).

@@

\emph{FIXME: Maybe better clarify what `existing data' in the next sentence means. I think you are referring to data where we have to evaluate out-of-sample rather than with the `true' effects in-sample, right?}

@@

\emph{README: The conclusion not to consider SEM trees which performed rather well in the simulation studies is a bit unsatisfying. Also, longRPart was not included in the simulation studies. Overall, the new insights gained compared to the simulation studies are not so huge. So maybe we should discuss again how to best present this?}

@@

The ECLS-K datasets have exceptionally large sample sizes. We therefore employed random sampling to obtain training samples of $N=250$ and $N=1,000$ children, likely more representative of real-world studies in psychology. For each combination of response variable (math, reading, or science) and sample size, we performed 100 repetitions. We evaluated predictive accuracy by computing the mean squared difference between predicted and observed response variable values for all children not included in the training sample of the current repetition. Note that this measures the accuracy of the fixed-effects predictions only. Tree size was measured by the number of splits in each tree, and computation time was measured in seconds.

@@

## Ideally, would refit all longRPart_s trees which had convergence errors
## in one of the nodes. And then use NA for prediction, or refit a model
## without random slopes. This might paint a slightly difference picture.
## However, I do not expect dramatic differences, given that results
## for longRPart and longRPart_s are quite similar, and seem to differ
## most for science, which was most strongly affected by the mistake.

@@

LongRPart trees comprising both random intercepts and slopes yielded convergence errors in \Sexpr{(table(is.na(math_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the math data and in \Sexpr{(table(is.na(scie_MSEs$longRPart_s))["TRUE"]/200)*100}\% of samples from the science data. All other model-fitting attempts converged successfully. Results are presented for successful model estimations.

@@

Table~\ref{tab:application_MSEs} presents means and standard deviations of the MSEs. Only few differences in performance between LM(M) and longRPart trees are larger $> 1$ standard error: For reading with $N=250$ and for science with both sample sizes. Note that the standard errors in the pairwise $t$-tests get smaller with a larger number of replications, while the standard errors in Table~\ref{tab:application_MSEs} do not. The latter thus provide a more useful measure of effect size, while the former have higher power to detect differences between methods.

Table~\ref{tab:application_MSEs} indicates that LMM trees estimating only a random intercept and employing cluster-level covariances performed best for predicting math and reading abilities. For predicting science abilities, LMM trees initializing estimation with the random effects and estimating only a random intercept performed best. Random-effects initialization was the worst-performing LMM tree approach for math and reading.

For longRPart, trees comprising both a random intercept and slope performed best for predicting math and reading, trees with a random intercept only performed best for predicting science. The differences in performance between math and reading on the one hand and science can perhaps be explained by the lower number of repeated assessments of science abilities.

@@

\begin{figure}[!bt]
\caption{Tree sizes for trees fitted to math (top), reading (middle), and science (bottom) ability trajectories.}
\begin{subfigure}{1.2\textwidth}
<<echo=FALSE, fig=TRUE, height=2, width=8>>=
## Tree size
load("sizes, N = 250, Math.Rda")
math_sizes <- cbind(sizes, N = 250)
load("sizes, N = 1000, Math.Rda")
math_sizes <- rbind(math_sizes, cbind(sizes, N = 1000))
#sapply(math_sizes, function(x) table(is.na(x)))
#sapply(math_sizes, function(x) tapply(x, math_sizes$N, mean, na.rm =TRUE))

@@

Figure~\ref{fig:application_sizes} presents tree size distributions for each of the partitioning approaches. Interestingly, whereas LM(M) trees implement a larger number of splits with increasing sample size, this pattern is reversed for longRPart trees. For math and reading, the best-performing LM(M) tree approach in terms of predictive accuracy yields a smaller number of splits than the best-performing longRPart approach when sample size is small. This pattern reverses with large sample size, where the best-performing LM(M) tree approach yields a larger number of splits than the best-performing longRPart approach. For the math data, the best-performing longRPart approach yields a smaller number of splits than the best-performing LM(M) tree approach.
